---
title: "Introduction to GIS in R"
author: "Where do we go from here?"
date: '[contact@appliedepi.org](mailto:contact@appliedepi.org)'
output:
  xaringan::moon_reader:
    seal: TRUE
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: [xaringan-themer.css] 
---

```{r, eval=F, echo=F, include = F}
# Must do in order to render.

pacman::p_load(xaringan)
devtools::install_github("gadenbuie/xaringanExtra")
devtools::install_github("gadenbuie/countdown")
remotes::install_github("mitchelloharawild/icons")
icons::download_fontawesome()

# Render with xaringan::infinite_moon_reader()
# Slides will appear in viewer, and will update as you edit/save
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3  ## retina more effective than w/h (according to twitter)
                      # fig.width = 16, 
                      # fig.height = 10
                      )
## I dont know why this is included in the example xaringan slide 
## but is something to do with background images
options(htmltools.dir.version = FALSE)

## install and load necessary packages 
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  tidyverse,  # data management and visualization
  gtsummary,  # summary tables
  flair,      # coloring text
  kableExtra, # for output tables
  xaringanthemer,  # for styling presentation 
  sf,
  ggspatial
)

library(countdown)
```


```{r  xaringan-themer, include = FALSE}

## define presentation colours (theme) using {xaringanthemer} package 
## https://pkg.garrickadenbuie.com/xaringanthemer/articles/xaringanthemer.html

## epirhandbook logo colours: 
  ## blue: "#00538c"
  ## green: "#007732"
  ## lighter green: "#48a878"

## see ?style_mono_accent for all the things can customise
style_mono_accent(
  base_color = "#00538c", 
  link_color = "#48a878", 
  # add logo to the title page (bit bigger)
  title_slide_background_image = "https://raw.githubusercontent.com/appliedepi/slides/master/images/logo.png",
  title_slide_background_position = "95% 95%",
  title_slide_background_size = "25%",
  ## add logo to all following slidesw
  background_image = "https://raw.githubusercontent.com/appliedepi/slides/master/images/logo.png", 
  background_size = "10%",
  background_position = "100% 0%"
)
```

```{css, echo=F}
    .remark-slide table{
      border: none
    }
    .remark-slide-table {
      
    }
    tr:first-child {
      border-top: none;
  }
    tr:last-child {
    border-bottom: none;
  }
```


```{r, echo=F, eval=T}
surv_raw <- rio::import(here::here("data", "surveillance_linelist_20141201.csv"))

surv <- rio::import(here::here("data", "surveillance_linelist_clean_20141201.rds"))
```



# Course recap 

So far we have used multiple data sources to learn about: 

* Shapefiles, {sf}, and CRS
* Spatial joins and admin levels
* {ggplot2}, basemaps, and bounding boxes
* Choropleth maps and heat maps
* Spatial analysis including NN, buffer analysis
* Interactive plots with {leaflet}

---

# Shapefiles and CRS


In future analyses, we need to think about: 

Where to source shape files from? 
  - [Open Street Map](https://www.openstreetmap.org/#map=5/-28.153/133.275)
  - [Natural Earth Data](https://www.naturalearthdata.com/downloads/)
  - [Esri Open Data Hub](https://hub.arcgis.com/search)
  - [NASA's Socioeconomic Data and Applications Center](https://sedac.ciesin.columbia.edu/)
  - There are **many more** resources not listed here!

--

Different regions use different types of CRS
  - E.g. Sierra Leone vs. Fulton county
    - Did they use the same format? 
  - What does your work region use? **Look it up!** 

---

# Importance of admin levels

We must always consider which administrative level to use: 
- admin1, admin2, admin3, country, district, zip codes, etc. 

--

How can we join our survey data and our shape files? 
 - `st_joins`, `st_within`, and many more options available

--

At what stage do we format our survey data to class `sf`?
 - Before joining with shape files or after?

--

How do we want our data formatted?
 - District or admin level choropleths? 
 - Points of cases using lat and lon


---
# Spatial analysis 

We have only touched the tip of the iceberg! There are many other more advanced methods. Some examples and tutorials available:  

Nearest Neighbour analysis using road networks 
   - See [this link](https://gis.stackexchange.com/questions/451410/doing-nearest-neighbors-accounting-for-road-distance-in-r) for advanced methods using roads

--

Interactive maps with leaflet
- [See this bookdown chapter on {leaflet} for reference](https://bookdown.org/ededeban/ConsBioMap/interactive.html)
- [Try out this practical to test your skills!](https://leafletjs.com/examples/choropleth/)

--

[More advanced spatial analysis beyond this course](https://medium.com/@milos.colic/scalable-spatial-nearest-neighbours-with-mosaic-336ce37edbae)
- Spatial autocorrelation
- Gravity models
- Neighbourhood-joining models

Many other methods out there! Your turn to explore!

---
# Resources available 

There are a multitude of open-source resources out there! 

[Applied Epi](https://appliedepi.org): 

- Applied Epi [community](https://community.appliedepi.org/)
- Applied Epi handbook [GIS chapter](https://epirhandbook.com/en/gis-basics.html)
  - We have covered most of this in the course but it is a good reference point
  - Includes plotting with {tmap}, another useful plotting package
- Applied Epi [support desk](https://support.appliedepi.org/)
   - Have your organisation reach out about purchasing **support desk access** for your team! 

--

Other resources:

- [Stack exchange](https://stackexchange.com/)
- [Stack overflow](https://stackoverflow.com/)
- [GIS handbook](https://www.spatialanalysisonline.com/An%20Introduction%20to%20Spatial%20Data%20Analysis%20in%20R.pdf)
- [Google](https://www.google.com/) is your friend! 

---
# Stay in touch! 

## We hope you enjoyed the course! 

Feel free to follow up with any questions or feedback: 

[contact@appliedepi.org](contact@appliedepi.org)


