---
title: "R Markdown demonstration"
subtitle: "Instructor guide"
output: html_document
theme: sandstone
date: 'Date revised: 2022-03-02'
---


```{r, echo=FALSE}
# Put logo in upper-right of document
htmltools::img(src = knitr::image_uri(here::here("images", "logo.png")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width = "200px",
               heigth = "200px")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Notes  

In this demonstration, you will show the students the power of R Markdown. Do not explain in too much detail - they will get detail in the subsequent slides and in their exercise. But show them the breadth of Rmd and how quickly they can create pretty (or editable) outputs. 

# Preparation  

0. Ensure that you have `tinytex` installed, Microsoft Word installed, pandoc is findable, your have {rmarkdown} installed, and you **test the Rmd scripts** before the live demo.  
1. Clear your desktop of clutter.  
2. Turn off pop-up notifications from desktop versions of WhatsApp, Signal, Teams, etc.  
3. Open RStudio and adjust the upper-right R project drop-down menu to close any open projects. This way, when you open RStudio it opens as "Project: None".  
5. Open the R script that was written to start this demonstration 
6. Have this file open to follow (or better: print it and have on your desk)

# Setup  

1. **In Google Meets**:  
     * If you have two displays: share your second display, which contains RStudio and the file explorer. The first display will be contain the Meet.  
     * If you only have one display, show your entire screen and ask an assistant instructor for help with any questions/hands that you cannot see. (TODO maybe there is a way to avoid mirroring in Meets?)  

</br>

1. **Open the R script "demo_rmd_guide.R"** from within the intro_course R project  
     * (1-2min) Begin showing the sections of this script. They should be familiar. Explain that it is a simplified version of the participants script.  
     * Run the script line-by-line to show the outputs (tables, epicurve, age pyramid, etc.)  
     
2. **Open a new R Markdown script**  
     * Name it "R Markdown demonstration"  
     * Explain that this is an R Markdown script, and that it can be converted into a document (e.g. Word, PDF)  
     * Explain that there are two sections - writing, and code.
     
3. **knit the default document to Microsoft Word**  
     * Point out the "Render" RStudio pane that shows the progress of the "rendering"  
     * Show that you can edit the Microsoft Word document  
     
4. **knit the default document to PDF**  
     * Explain that PDF can be more difficult, but is useful. Remind them that they can always use Word and then save as PDF.  

5. **knit the default document to HTML**  
     * Explain that an HTML file allows much more interactivity
     
6. **Delete the default content and transfer the R script into the Rmd**  
     * Title (to YAML)
     * Author (create in YAML)  
     * Subtitle (create in YAML)  
     * Note date in YAML  
     
     * Load packages  
     * Define parameters  
     * Import data  
     * Clean data  
     * Descriptive table  
     * Plots...  
     
7. **Begin to convert to an organized Rmd**  
     * Begin with adding Headers (e.g. # Descriptive tables,  # Plots, and ## Epicurve)  
     * Then add some intro text at the top: "This is a report on the Ebola outbreak in Sierra Leone. The current date is {r Sys.Date()}... etc.  
     * Demonstrate writing in normal text, 
     * Show how to make bullets  
     * Show some **bold text** and some *italics* and "Now I am going to write the function `mutate()` in code-text.  
     

     
8. **Re-render as HTML**  

9. **Adjust the report parameters**  
     * Adjust to `echo = FALSE, warning = FALSE, message = FALSE` and re-render as HTML  
     * Set the HTML YAML to below and re-render:  

```{r, echo=T, eval=F, results='asis'}
output:
  html_document:
    toc: true
    toc_depth: 2  

```
     
Now adjust it so the TOC is floating, and re-render:  

```{r, echo=T, eval=F, results='asis'}
output:
  html_document:
    toc: true
    toc_depth: 2  
    toc_float: true
    number_sections: true
```

Now edit the # Plots header to read: # Plots {.tabset} (ensure the sub-headings under are ## double-headers)  

Re-render


End (go to slides)
