---
title: "Introduction to GIS in R"
author: "COVID-19 GIS case study"
date: '[contact@appliedepi.org](mailto:contact@appliedepi.org)'
output:
  xaringan::moon_reader:
    seal: TRUE
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: [xaringan-themer.css] 
---

```{r, eval=F, echo=F, include = F}
# Must do in order to render.

pacman::p_load(xaringan)
devtools::install_github("gadenbuie/xaringanExtra")
devtools::install_github("gadenbuie/countdown")
remotes::install_github("mitchelloharawild/icons")
icons::download_fontawesome()

# Render with xaringan::infinite_moon_reader()
# Slides will appear in viewer, and will update as you edit/save
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3  ## retina more effective than w/h (according to twitter)
                      # fig.width = 16, 
                      # fig.height = 10
                      )
## I dont know why this is included in the example xaringan slide 
## but is something to do with background images
options(htmltools.dir.version = FALSE)

## install and load necessary packages 
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  tidyverse,  # data management and visualization
  gtsummary,  # summary tables
  flair,      # coloring text
  kableExtra, # for output tables
  xaringanthemer,  # for styling presentation 
  sf,
  ggspatial
)

library(countdown)
```


```{r  xaringan-themer, include = FALSE}

## define presentation colours (theme) using {xaringanthemer} package 
## https://pkg.garrickadenbuie.com/xaringanthemer/articles/xaringanthemer.html

## epirhandbook logo colours: 
  ## blue: "#00538c"
  ## green: "#007732"
  ## lighter green: "#48a878"

## see ?style_mono_accent for all the things can customise
style_mono_accent(
  base_color = "#00538c", 
  link_color = "#48a878", 
  # add logo to the title page (bit bigger)
  title_slide_background_image = "https://github.com/appliedepi/intro_course/raw/main/images/logo.png",
  title_slide_background_position = "95% 95%",
  title_slide_background_size = "25%",
  ## add logo to all following slidesw
  background_image = "https://github.com/appliedepi/intro_course/raw/main/images/logo.png", 
  background_size = "10%",
  background_position = "100% 0%"
)
```

```{css, echo=F}
    .remark-slide table{
      border: none
    }
    .remark-slide-table {
      
    }
    tr:first-child {
      border-top: none;
  }
    tr:last-child {
    border-bottom: none;
  }
```


```{r, echo=F, eval=T}
surv_raw <- rio::import(here::here("data", "surveillance_linelist_20141201.csv"))

surv <- rio::import(here::here("data", "surveillance_linelist_clean_20141201.rds"))
```


---
# Data used in this course  

So far we have used the Ebola case study data

* Ebola case study in Sierra Leone
     * **Surveillance linelist** (primary dataset)  
     * Hospital linelists from 5 hospitals  
     * Population data stratified by demographic groups and admin levels
     * Sierra Leone shape files stratified by admin levels
     * Health facility shape files

Now we introduce a new case study: 

* COVID-19 case study in USA
     * **Surveillance linelist** (primary dataset)
     * Population data stratified by cities
     * Fulton County, Georgia, USA, shape files
     * Fulton County, Georgia, USA, map tiles
     
.footnote[All data are not real. All files can be found in the course folder.]  


---

# First steps in a new analysis

As we are using new data which we have never seen before, it is useful to explore the data first
- Take a look at which columns are available in our **surveillance linelist** 
- What geographpic levels (e.g. adm, lat, lon) do we have available in our survey data?
- What geographic boundaries are available in our shape files? 
- **Note** we have population data by city
  - How would we join our surveillance linelist data to our population data? 
  - How would we join our surveillance linelist dat to our shape files? 
  - What can we use our population data to calculate?

---
# Cultural notes: *Zip* codes

If you are not familiar with the hierarchy of government jurisdictions in the USA:

- Georgia (state) contains over 150 different counties
  - Counties can contain multiple cities and/or municipalities
- There are over 500 cities within the state of Georgie
  - **Most** cities fall within one county, but some span across multiple
- *ZIP* codes are the US postal code system. 
  - *ZIP* codes can span across cities and counties
- Fulton county spans across **25** ZIP codes

---

# Cultural notes: *Race vs. ethnicity* 

In the USA, census data often include data on *Race* and *Ethnicity* as two separate columns. 

Ethnicity options are typically:  
* "Hispanic or Latino" or,
* "Not Hispanic or Latino"  

Race options often include:  

* "American Indian or Alaska Native" (sometime abbreviated as AI/AN)  
* "Asian"
* "Black or African American"
* "Native Hawaiian or Other Pacific Islander"
* "White"  


---

# Surveillance periods 

We also might want to think about viewing *case counts over time* using **surveillance periods** 

Why might this be useful? 

- Compare *incidence* across **time periods** (e.g. week to week)
- Observe how and where the outbreak is growing fastest
- How might we *create surveillance periods* in R for our analysis? 
   - `floor_date()` from {lubridate} 


---

# Surveillance periods

In R, we can set up surveillance periods easily based on our data. 

```{r, echo = TRUE, tidy = FALSE}

# create a date object with most relevant date of complete data
# minus 5 days to account for lag in reporting lab results

# first day of complete data (accounting for lab reporting delays)
surveillance_date <- as.Date("2021-06-29") - 5

# create an object to mark start of latest complete week of data 
# floor_date() rounds down to the closest week here
surveillance_week_start <- floor_date(
  # subtract 7 days from surveillance_date
  surveillance_date - 7,
  # round down to nearest week start date
  unit = "week", 
  # define week to start on Sunday (typical in USA)
  week_start = 7)


# define recent (past 14 days) and previous (28 to 14 days prior)
recent_period   <- seq(surveillance_week_start  - 13, surveillance_week_start, by = 1)
previous_period <- seq(surveillance_week_start  - 27, surveillance_week_start - 14, by = 1)
```


How could we set up `surveillance_date` in RMarkdown? Think about the *params* key in *YAML* section. 

---

# Data cleaning and analyses

*NOTE* Data cleaning is not the focal point of this course
- Data cleaning pipeline provided in the exercsies to copy and paste 
- Feel free to attempt on your own first, but don't spend too much time on it

The focal point of this session is to apply our newly learned skills to new data. Things to consider: 

- What *class* is our survey analyses object? What *class* does it need to be for GIS? 
- Which columns can you use for transforming data to *class* `sf`? 
- What is the CRS of our shapefile? Does this align with our case linelist? 
- Which `st` join function should we use to join linelist, population, and shape files?

---

class: inverse, center, middle

## Demo & Exercise

---
