---
title: "Module 1: R Setup, syntax, data import"
subtitle: "Instructor guide"
output: html_document
theme: sandstone
date: 'Date revised: 2022-03-02'
---


```{r, echo=FALSE}
# Put logo in upper-right of document
htmltools::img(src = knitr::image_uri(here::here("images", "logo.png")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width = "200px",
               heigth = "200px")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Notes  

This module gets the students setup in RStudio with a project, demo script, and interacting with the learnr exercises.  

At the beginning, you will demonstrate how to open a learnr exercise. Then, you can let the students work through the exercise and assist them as-needed.  

# Preparation  

1. Clear your desktop of clutter.  
2. Turn off pop-up notifications from desktop versions of WhatsApp, Signal, Teams, etc.  
3. Open RStudio and adjust the upper-right R project drop-down menu to close any open projects. This way, when you open RStudio it opens as "Project: None".  
4. Open a file explorer window, showing the location where you will save the R project, and have it ready to show.  
5. Open this file (or better: print it and have on your desk)

# Setup  

1. **In Google Meets**:  
     * If you have two displays: share your second display, which contains RStudio and the file explorer. The first display will be contain the Meet.  
     * If you only have one display, show your entire screen and ask an assistant instructor for help with any questions/hands that you cannot see. (TODO maybe there is a way to avoid mirroring in Meets?)  

</br>

# Introduction and welcome  

Welcome the group and introduce yourself. 

Explain how the breakout sessions will work:  

* You are available to answer any questions and help people  
* The primary activity is them working through the interactive exercise for this module.  
* If there is something that they do not understand, probably there are others that do not understand as well.  
* They can turn their camera off while working, if they prefer  
* The breakout will reconvene towards the end of the breakout time to discuss and ensure that everyone was successful.  


# Initial setup 

Explain that there is some initial setup that needs to be done.  

**Give them the link to the course website and tell them to follow the "Initial setup instructions".**

When they reach the part about downloading our Applied Epi R package, they will be prompted to run the following commands:  

`remotes::install_github("appliedepi/introexercises", dep = TRUE, force = TRUE)`  

  * If you see a warning about needing RTools, do not worry. This is simply a warning message, but the install should proceed without error.  
  * If prompted to update some R packages, select "All" to update all of the packages.  
  * If you continue to be prompted for package updates, you can select "None" and continue without further updates.  

When they open the tutorials pane for the first time, they may need to adjust the View size of RStudio, and adjust the pane edges in order to see the sidebar.  

If they pop-out the tutorial into a separate window, they cannot re-insert it into RStudio. In this case, just have them close the pop-out window and re-open the tutorial. Their work in the tutorial (quiz questions, etc.) will be saved.  


# Tutorial 1

The students should then proceed into Tutorial 1, which is called "Applied Epi - R setup, syntax, data import.  



# Finish
Close RStudio.    
Return to begin the plenary debrief with the entire class.  


