---
title: "Introduction to R for<br>Applied Epidemiology"
subtitle: "COVID-19 case study"
author: "March 2022"
date: '[contact@appliedepi.org](mailto:contact@appliedepi.org)'
output:
  xaringan::moon_reader:
    seal: TRUE
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: xaringan-themer.css
---

```{r, eval=F, echo=F, include=F}
# Must do in order to render.

pacman::p_load(xaringan)
devtools::install_github("gadenbuie/xaringanExtra")
remotes::install_github("mitchelloharawild/icons")
icons::download_fontawesome()

# Render with xaringan::infinite_moon_reader()
# Slides will appear in viewer, and will update as you edit/save
```


```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3  ## retina more effective than w/h (according to twitter)
                      # fig.width = 16, 
                      # fig.height = 10
                      )
## I dont know why this is included in the example xaringan slide 
## but is something to do with background images
options(htmltools.dir.version = FALSE)

## install and load necessary packages 
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  sf,         # spatial
  ggspatial,  # spatial
  kableExtra, # for output tables
  xaringanthemer,  # for styling presentation
  tidyverse  # data management and visualization

)

## load packages from github
pacman::p_load_gh(
     "R4IDSR/epichecks"   # bivariate colour maps 
)


# data prep --------------------------------------------------------------------
surv <- rio::import(here::here("data", "surveillance_linelist_clean_20141201.rds"))
```

```{r  xaringan-themer, include = FALSE}

## define presentation colours (theme) using {xaringanthemer} package 
## https://pkg.garrickadenbuie.com/xaringanthemer/articles/xaringanthemer.html

## epirhandbook logo colours: 
  ## blue: "#00538c"
  ## green: "#007732"
  ## lighter green: "#48a878"

## see ?style_mono_accent for all the things can customise
style_mono_accent(
  base_color = "#00538c", 
  link_color = "#48a878", 
  ## add logo to the title page (bit bigger)
  title_slide_background_image = "https://github.com/appliedepi/intro_course/raw/main/images/logo.png", 
  title_slide_background_position = "95% 95%", 
  title_slide_background_size = "25%", 
  ## add logo to all following slides
  background_image = "https://github.com/appliedepi/intro_course/raw/main/images/logo.png", 
  background_size = "10%",
  background_position = "100% 0%"
)
```

```{css, echo=F}
    .remark-slide table{
      border: none
    }
    .remark-slide-table {
      
    }
    tr:first-child {
      border-top: none;
  }
    tr:last-child {
    border-bottom: none;
  }
```


# Today: objectives & schedule  

**In this module we aim to help you:**  

* Apply your learning to a new dataset, without code solutions readily available  

--

* See outputs and replicate them using R  

--

* Write original R code that is understandable by someone else  

--

* Review another person's original code  

--

* Complete an end-course feedback survey  

```{r, echo=FALSE, warning=F, message=F}
outline <- dplyr::tribble(
  ~Time, ~Location, ~Topic,
  "15 minutes",     "Main",      "Introduce the COVID-19 case study",
  "2 hours",        "Individual","COVID-19 case study",
  "45 minutes",     "Main",      "Partner code review",
  "5 minutes",      "Main",      "Closing",
  "15 minutes",     "Main",      "Feedback survey"
)

outline %>% 
  flextable::qflextable() %>% 
  flextable::add_footer_lines("Breaks are incorporated above, but not shown as rows")
```


---

# COVID-19 report

Fulton County, GA

You will have 2 hours to replicate the report  


---

# COVID-19 Data

Scrambled and anonymized  

- Linelist  
- City populations in Fulton County  


---

# Support available  

- Instructors will be available in breakouts room  
- Each will handle different parts of the report, so go to the room that most fits your question  


---
# Have fun  

This is not a test. Do the best you can. Enjoy reviewing about all that you have learned.  


---
# Code review  

When we re-convene, we will split into partners (randomly-assigned) and you can review the other person's code.  

We will provide a brief guide on how to review code.  


---
# Instructor feedback  

Our instructors will review your work after the course and provide you feedback by email.  

