---
title: "Applied Epi"
subtitle: "Taking control of your data"
format:
  revealjs: 
    code-line-numbers: false
    slide-number: true
    chalkboard: 
      buttons: true
    controls: false
    multiplex: false
    preview-links: auto
    logo: ../../images/Applied_Epi_logo.png
    css: ../../images/styles.css
    footer: <https://appliedepi.org>
resources:
  - demo.pdf
fig-align: center
editor: 
  markdown: 
    wrap: 72
---

```{r}

# Packages
pacman::p_load(rio, here, janitor, dplyr, stringr, lubridate, gtExtras, gt, tidyr, ggplot2, glue, datapasta, reprex, flextable, gtsummary)


# Lab data -------------------------------
# Labs
df_lab_luh <- import(here("intror/gi_2021-06-17_lab_luh.csv"))
df_lab_vnrl <- import(here("intror/gi_2021-06-17_lab_vnrl.csv"))
df_lab_ghc <- import(here("intror/gi_2021-06-17_lab_ghc.csv"))


# Load data as needed 
df_raw <- import(here("intror/gi_2021-06-17_linelist.xlsx")) 

# Other symptoms
df_symptoms <- import(here("intror/gi_2021-06-17_symptoms.xlsx")) 

# Clean linelist data ---------------------------

df <- df_raw |>

  # Clean column names
  clean_names() |>
  rename(date_onset = symptoms_started) |>

  # Change character class
  mutate(admission = mdy(admission),
         date_onset = mdy(date_onset),
         discovered = dmy(discovered),
         found_on = mdy(found_on),
         reported = ydm(reported),
         told_on = mdy(told_on),
         report  = ymd(report)) |>

  # Clean column content
  mutate(region = str_to_title(region),
         sex = str_to_title(sex),
         sex = case_match(sex,
                          c("F", "Fem", "Female") ~ "Female",
                          c("M","Man", "Male") ~ "Male",
                          .default = NA),
         region = case_match(region,
                             c("Bexar") ~ "Bexar",
                             c("Keda", "Kedah") ~ "Keda",
                             c("Liora", "Lioraah") ~ "Liora",
                             c("Nellus", "Nelos") ~ "Nelos",
                             c("New Ardin", "New Arden") ~ "New Arden",
                             c("Sorel", "Sorell") ~ "Sorel",
                             c("Taris") ~ "Taris",
                             c("Varek") ~ "Varek",
                             c("West Morin") ~ "West Morin",
                             .default = NA_character_))  |>

  # Deduplicate to one row per case, based on case ID
  distinct(id, .keep_all = TRUE) |>

  # Bring data saved across multiple columns together
  mutate(date_report =  coalesce(told_on, reported, discovered, found_on, report)) |>
  select(-told_on, -reported, -discovered, -found_on, -report) |>

  # Separate outcome and date (tidyr)
  separate_wider_delim(reported_outcome,
                       delim = ":",
                       names = c("outcome", "date_outcome")) |>
  mutate(date_outcome = ymd(date_outcome)) |>

  # Create date of birth and age
  unite("date_of_birth",
        c("birthyear", "month_of_birth", "daybirth"), sep = "-") |>
 # mutate(date_of_birth = glue(birthyear-{month_of_birth}-{daybirth}))
  mutate(date_of_birth = ymd(date_of_birth)) |>

  # Calculate age
  mutate(age_days = date_report - date_of_birth,
         age_years = age_days / 365.25,
         age = floor(age_years),
         age = as.numeric(age)) |>

  # Create columns labelling diarrhoea or dehydration
  mutate(symptoms = str_to_lower(symptoms),
         symp_diarrhea = str_detect(symptoms, "diarrhoea|diarrhea|watery stool|loose stool|awd"),
         symp_dehydration = str_detect(symptoms, "dehydra|lack of fluids|lack fluids|dry mouth")) |>

  # Create case definition column
  mutate(case_def = case_when(cholera_rdt_positive==1 ~ "Probable",
                              symp_diarrhea ==TRUE & symp_dehydration==TRUE ~ "Suspected",
                              TRUE ~ "Unclear")) |>

  mutate(case_def = factor(case_def, levels = c("Unclear", "Suspected", "Probable"))) |>

  # Create week column for onset date
  mutate(date_onset_week = floor_date(date_onset, "week")) |>

  # Create week column for report date
  mutate(date_report_week = floor_date(date_report, "week")) |>

  # Create column for whether cases are new
  mutate(date_report_7 = if_else(date_report >=  ymd("2021-06-17")-7, "new", "previous"))

# Clean and append lab data------------------------

df_lab_luh <- df_lab_luh |>
  clean_names()

df_lab_ghc <- df_lab_ghc |>
  rename(cholera_pcr_result = pcr_result,
         cholera_pcr_positive = pcr_positive,
         cholera_pcr_date_test = pcr_date_test,
         cholera_pcr_date_result = pcr_date_result)

df_lab <- bind_rows(df_lab_luh, df_lab_ghc, df_lab_vnrl)

# Link datasets  -------------------------------------

df_joined <- df |>
  left_join(df_lab, by = "id") |>
  left_join(df_symptoms, by = "id")

df_joined <- df_joined |>
  mutate(case_def = case_when(cholera_pcr_positive==1 ~ "Confirmed",
                              cholera_rdt_positive==1 ~ "Probable",
                              symp_diarrhea ==TRUE & symp_dehydration==TRUE ~ "Suspected",
                              TRUE ~ "Unclear")) |>
  mutate(case_def = factor(case_def, levels = c("Unclear", "Suspected", "Probable", "Confirmed")))



```

# Module 8: Part B

## Scenario {.exercise-slide}

::: incremental
-   You have decided to collate all your analyses in an automated report, which will require formatted tables
-   Of course, you will format your tables in R!

:::

## Module 8 learning objectives

In Part A you:

-   Learned how to reshape data with `pivot_wider()` and pivot_longer()\`
-   Practiced making summary tables with pivoted data

In Part B you will:

-   Learn to create formatted tables with {janitor}, {flextable}, and {gtsummary}

# Tables for presentation

## Tables for presentation

You've learned how to create summary tables in R using:

- *tabyl()* for quick tabulations
- *group_by()* and *summarise()* for showing several metrics across groups
- *pivot_wider()* to reshape data for presentation
- Even *_join() to link tables

Now, let's learn how to format these tables for automated reporting!


# {janitor}

## adorn_*() functions

::: slide-header
{janitor}
:::

`tabyl()` from {janitor} is fantastic for quick tabulations, but is ugly. It can be customized with a range of adorn_*() functions.

:::::: columns
::: {.column width="48%"}
**For example, instead of just this: **

::: {.code-box-small style="font-size: 40px;"}
```{r eval=F, echo=T}
df_joined |> tabyl(case_def)
```
:::

::: arrow-box
<br>
↓
:::

::: output-box
```{r}
df_joined |> tabyl(case_def) |> 
  gt::gt() |>  gtExtras::gt_theme_guardian() |> 
  gt::tab_options(table.font.size = 17) 
```
:::
::::::

::: {.column width="2%"}
:::

::: {.column width="48%"}
**... you can add totals and formatting!**

::: {.code-box-small style="font-size: 40px;"}
```{r eval=F, echo=T}
df_joined |> tabyl(case_def) |> 
  adorn_totals() |> 
  adorn_pct_formatting()
```
:::

::: arrow-box
↓
:::

::: output-box
```{r}
df_joined |> tabyl(case_def) |> 
  adorn_totals() |> 
  adorn_pct_formatting() |> 
  gt::gt() |>  gtExtras::gt_theme_guardian() |> 
  gt::tab_options(table.font.size = 17) 
```
:::

:::::

:::::::

## adorn_*() functions

::: slide-header
{janitor}
:::

**Here are other adorn_* functions:**

::: {#general-box style="text-align: left; width: 1050px; background: #ebebeb; margin: 0 auto; font-size: 25px;"}
| **Function** | **Purpose** |
|-------------------------|-----------------------------------------------|
| **`adorn_totals()`** | Adds row and column totals |
| **`adorn_percentages()`** | Converts counts to percentages |
| **`adorn_ns()`** | Displays both counts and percentages (typically used **after** `adorn_percentages()` |
| **`adorn_pct_formatting()`** | Formats percentages (e.g., rounding, adding `%`) |
| **`adorn_rounding()`** | Rounds numeric values |
| **`adorn_title()`** | Adjusts column names  |
:::

::: {.obs-box .fragment}
💡 **NOTE** For brevity we do not describe all function details and arguments. Instead, look them up by typing `?function_name` (e.g. `?adorn_totals`) in your R console!
:::


# {flextable}

## {flextable}

::: slide-header
{janitor}
:::

For publication-ready tables that can go straight into automated
reports, we recommend {flextable}.

-   *Purpose*: Create and customize formatted tables to output as images
    or into Word, PowerPoint, or HTML reports.
-   *Why*: This package provides good control over styling, layout, and
    formatting
-   *How*: **Create a summary table first**, then apply the
    `flextable()` function. You can then apply several functions from
    the {flextable} package to further customize.

::: code-box
```{r, eval=F, echo=T}
flextable(summary_table) 
```
:::

## {flextable}

::: slide-header
{janitor}
:::

:::::: columns
::: {.column width="48%"}
This is what printing the simple case category table looks
like in your R console: 

::: {.code-box-small style="font-size: 40px;"}

```{r eval=F, echo=T}
df_joined |> tabyl(case_def) |> 
  adorn_totals() |> 
  adorn_pct_formatting()
```
:::

::: arrow-box
↓
:::

::: {.output-box style="font-size: 40px;"}
```{r}
df_joined |> tabyl(case_def) |> 
  adorn_totals() |> 
  adorn_pct_formatting()
```
:::
::::::

::: {.column width="2%"}
:::

::: {.column width="48%"}
If you pipe into a `flextable()` it appears with minimal formatting in your Viewer: 

::: {.code-box-small style="font-size: 40px;"}
```{r eval=F, echo=T}
df_joined |> tabyl(case_def) |> 
  adorn_totals() |> 
  adorn_pct_formatting() |> 
  flextable()
```
:::

::: arrow-box
↓
:::

::: {.output-box style="font-size: 40px;"}
```{r}
df_joined |> tabyl(case_def) |> 
  adorn_totals() |> 
  adorn_pct_formatting() |> 
  flextable() |> 
  bg(bg = "white", part = "all") |> 
  fontsize(size = 19, part = "all")  |> 
  autofit()
```
:::

:::::

:::::::

## {flextable}

::: slide-header
{janitor}
:::

You can apply further formatting functions from the {flextable} package, e.g.:

::::::::: columns
:::: {.column width="53%"}
::: {.code-box-small style="width: 550px; font-size: 35px;"}
```{r echo=T, eval=F}
df_joined |> tabyl(case_def) |> 
  adorn_totals() |> 
  adorn_pct_formatting() |> 
  flextable() |> 
  
  # Change column names in flextable
  set_header_labels(case_def = "Case category",
                    n = "Count",
                    percent = "Percent") |> 
  
  # Bold header and total (5th) row
  bold(part = "header") |> 
  bold(part = "body", i = 5) |> 
  
  # Add gray background to header
  bg(part = "header", bg = "gray90")

```
:::
::::

:::: {.column width="2%"}
::: arrow-box
→
:::
::::

:::: {.column width="40%"}
::: output-box
```{r}
df_joined |> tabyl(case_def) |> 
  adorn_totals() |> 
  adorn_pct_formatting() |> 
  flextable() |> 
  
  # Change column names in flextable
  set_header_labels(case_def = "Case category",
                    n = "Count",
                    percent = "Percent") |> 
  
  # Bold header and total (5th) row
  bold(part = "header") |> 
  bold(part = "body", i = 5) |> 
  
  # Add gray background to header
  bg(part = "header", bg = "gray90") |> 
  bg(part = "body", bg = "white") |> 
  fontsize(part = "all", size=18)


```
:::
::::
:::::::::

::: {.obs-box .fragment}
💡 **NOTE** Don't worry about understanding all these functions immediately, we're just showing you what is possible!
:::

## {flextable}

::: slide-header
{janitor}
:::

**Here are some {flextable} functions for formatting your table:**

::: {#general-box style="text-align: left; width: 1000px; background: #ebebeb; margin: 0 auto; font-size: 25px;"}

| **Function**       | **Purpose**                                         |
|--------------------|-----------------------------------------------------|
| `bold()`           | Makes text bold in the table                        |
| `color()`          | Changes text color                                  |
| `set_caption()`    | Adds a caption to the table                         |
| `merge_v()`        | Merges cells vertically in a column                 |
| `align()`          | Aligns text (left, center, right)                   |
| `add_header_row()` | Adds a new header row                               |
| `border_inner()`   | Adds inner borders between cells                    |
| `theme_*()`        | Applies predefined themes (e.g., `theme_vanilla()`) |

:::

::: {.obs-box .fragment}
💡 **NOTE**  Most of these functions use arguments to specify which part of the table to format: `part` (= "body", "header", or "all"), `i` (for table row), and `j` (for column). <br>
Again, type `?function_name` in R for more information!
:::

# {gtsummary}

## tbl_summary() 

::: slide-header
{gtsummary}
:::

- *Purpose*: Summarizes several categorical and continuous variables in a dataset in a presentable format
- *Why*: Allows a quick look at the distributions of several columns at once
- *Example*: You want to prepare "Table 1" for a paper showing all the distribution of all demographic variables
- *How*: Apply `tbl_summary()` to a full linelist with only the columns of interest

::: code-box
```{r eval=F, echo=T}

data |> 
  select(col1, col2, col3) |> 
  tbl_summary()

```
:::

## tbl_summary()

See how you can quickly summarize several columns:

::::::::: columns
:::: {.column width="53%"}
::: {.code-box-small style="width: 550px; font-size: 35px;"}
```{r echo=T, eval=F}
df_joined |> 
  select(case_def, 
         outcome,
         age,
         symp_diarrhea) |> 
  tbl_summary()
```
:::

<br>

::: {.obs-box .fragment}
💡 **NOTE** See how R treats categorical and continuous variables differently, and specifies NAs as "Unknown"
:::
::::

:::: {.column width="2%"}
::: arrow-box
→
:::
::::

:::: {.column width="40%"}
::: output-box
```{r}
df_joined |> 
  select(case_def, 
         outcome,
         age,
         symp_diarrhea) |> 
  tbl_summary()
```
:::
::::
:::::::::


## tbl_summary()

Again, you can add further formatting: 

::::::::: columns
:::: {.column width="53%"}
::: {.code-box-small style="width: 550px; font-size: 35px;"}
```{r echo=T, eval=F}
df_joined |> 
  select(case_def, 
         outcome,
         age,
         symp_diarrhea) |> 
  tbl_summary()
```
:::
<br>

::: {.obs-box .fragment}
💡 **NOTE** This is an object of `gt` class, and <br> will also pop up in your Viewer. 
:::
::::

:::: {.column width="2%"}
::: arrow-box
→
:::
::::

:::: {.column width="40%"}
::: output-box
```{r}
df_joined |> 
  select(case_def, 
         outcome,
         age,
         symp_diarrhea) |> 
  tbl_summary(label = list(
              age ~ "Age", 
              outcome ~ "Outcome", 
              symp_diarrhea ~ "Reported diarrhea")) |> 
  bold_labels()   # Make variable names bold
```
:::
::::
:::::::::

# Exercise

## Exercise {.exercise-slide}

::: incremental
-   Your task is now to create presentable tables using the functions discussed today.

- Access exercises here: 

:::

# Recap

## 

## Your total code {.exercise-slide}

Your code should now look like this!

```{r eval=F, echo=T}

# Intro course
# Title of script: Cleaning and analysis of GI outbreak data
# Date started: 5th June 2021
# Name: Paula Blomquist

# Load packages-----------------------------

pacman::p_load(rio, here, janitor, dplyr, stringr, lubridate, tidyr, glue, ggplot2)

# Import data ------------------------------

# Linelist
df_raw <- import(here("data/gi_2021-06-17_linelist.xlsx"))

# Labs
df_lab_luh <- import(here("data/gi_2021-06-17_lab_luh.csv"))
df_lab_vnrl <- import(here("data/gi_2021-06-17_lab_vnrl.csv"))
df_lab_ghc <- import(here("data/gi_2021-06-17_lab_ghc.csv"))

# Other symptoms
df_symptoms <- import(here("data/gi_2021-06-17_symptoms.xlsx"))

# Clean linelist data ---------------------------

df <- df_raw |>
  
    # Clean column names
    clean_names() |>
    rename(date_onset = symptoms_started) |>
  
    # Change character class
    mutate(admission = mdy(admission),
           discovered = dmy(discovered),
           found_on = mdy(found_on),
           reported = ydm(reported),
           told_on = mdy(told_on),
           report  = mdy(report)) |>
  
    # Clean column content  
    mutate(region = str_to_title(region),
           sex = str_to_title(sex), 
           sex = case_match(sex, 
                            c("F", "Fem", "Female") ~ "Female",
                            c("M","Man", "Male") ~ "Male",
                            .default = NA),
           region = case_match(region,
             c("Bexar") ~ "Bexar",
             c("Keda", "Kedah") ~ "Keda",
             c("Liora", "Lioraah") ~ "Liora",
             c("Nellus", "Nelos") ~ "Nelos",
             c("New Ardin", "New Arden") ~ "New Arden",
             c("Sorel", "Sorell") ~ "Sorel",
             c("Taris") ~ "Taris",
             c("Varek") ~ "Varek",
             c("West Morin") ~ "West Morin",
             .default = NA_character_))  |> 
  
  # Deduplicate to one row per case, based on case ID
  distinct(id, .keep_all = TRUE) |>
  
  # Combine some columns
  unite("date_of_birth", 
          c("birthyear", "month_of_birth", "daybirth"), sep = "-") |> 
  mutate(date_of_birth = ymd(date_of_birth)) |>
  mutate(date_report =  coalesce(told_on, reported, discovered, found_on, report)) |> 
    select(-told_on, -reported, -discovered, -found_on, -report) |>
  
  # Create column for age
  mutate(age_years = as.numeric(floor((date_report -date_of_birth)/365.25))) |> 

  # Create columns labelling diarrhoea or dehydration    
  mutate(symptoms = str_to_lower(symptoms),
         symp_diarrhea = str_detect(symptoms, "diarrhoea|diarrhea|watery stool|loose stool|awd"),
         symp_dehydration = str_detect(symptoms, "dehydra|lack of fluids|lack fluids")) |>
  
  # Create case definition column
  mutate(case_def = case_when(cholera_rdt_positive==1 ~ "Probable",
                              symp_diarrhea ==TRUE & symp_dehydration==TRUE ~ "Suspected", 
                              TRUE ~ "Unclear")) |> 
  
  mutate(case_def = factor(case_def, levels = c("Unclear", "Suspected", "Probable"))) |> 
  
  # Create week column for onset date
 # mutate(date_onset_week = floor_date(date_onset, "week")) |> 
  
  # Create week column for report date
  mutate(date_report_week = floor_date(date_report, "week")) |> 

  # Create column for whether cases are new
  mutate(date_report_7 = if_else(date_report >=  ymd("2021-06-17")-7, "Reported in past 7 days", "Reported previously"))

# Clean and append lab data------------------------

df_lab_luh <- df_lab_luh |> 
  clean_names()

df_lab_ghc <- df_lab_ghc |> 
  rename(cholera_pcr_result = pcr_result,
         cholera_pcr_positive = pcr_positive,
         cholera_pcr_date_test = pcr_date_test,
         cholera_pcr_date_result = pcr_date_result)

df_lab <- bind_rows(df_lab_luh, df_lab_ghc, df_lab_vnrl)

# Link to df -------------------------------------

df_joined <- df |> 
  left_join(df_lab, by = "id")

df_joined <- df_joined |>   
  mutate(case_def = case_when(cholera_pcr_positive==1 ~ "Confirmed",
                              cholera_rdt_positive==1 ~ "Probable",
                              symp_diarrhea ==TRUE & symp_dehydration==TRUE ~ "Suspected", 
                              TRUE ~ "Unclear")) |> 
  mutate(case_def = factor(case_def, levels = c("Unclear", "Suspected", "Probable", "Confirmed"))) 


# Check of joins ---------------------------------

df_lab |> 
  tabyl(cholera_pcr_positive)

df_joined |> 
  tabyl(cholera_pcr_positive)

# Plots ------------------------------------------

# By report date and case definition breakdown
ggplot() +
  geom_bar(data = df_joined,
           aes(x=date_report_week,
               fill = case_def)) +
  scale_fill_manual(values = c("Unclear" = "gray70",
                               "Suspected" = "lightblue",
                               "Probable" = "seagreen",
                               "Confirmed" = "navy")) +
scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  labs(title = "Cases by report week, presented by case category and region",
       subtitle = glue("Data as of {max(df$date_report, na.rm=T)+1}"),
       fill = "Case category",
       x = "Week of report (week starting)",
       y = "Weekly case count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# By report date and case definition breakdown and region
ggplot() +
  geom_bar(data = df_joined,
           aes(x=date_report_week,
               fill = case_def)) +
  facet_wrap(.~region) +
  scale_fill_manual(values = c("Unclear" = "gray70",
                               "Suspected" = "lightblue",
                               "Probable" = "seagreen",
                               "Confirmed" = "navy")) +
scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  labs(title = "Cases by report week, presented by case category and region",
       subtitle = glue("Data as of {max(df$date_report, na.rm=T)+1}"),
       fill = "Case category",
       x = "Week of report (week starting)",
       y = "Weekly case count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
# Table of symptoms -----------------------------

# Pivot longer
df_symptoms_long <- df_joined |>
  pivot_longer(cols = starts_with("symp_"),
               values_to = "symptom_experienced",
               names_to = "symptom") |>
  select(id, case_def, region, symptom, symptom_experienced) |> 
  mutate(symptom = str_to_title(str_replace(symptom, "symp_", ""))) |> 
  mutate(symptom = str_replace(symptom, "_", " "))

# Simple table
table_symptoms <- df_symptoms_long |>
  group_by(symptom) |>
  summarize(reported = sum(symptom_experienced)) |>
  mutate(percent = scales::percent(reported/nrow(df_joined))) |>
  arrange(-reported)

# Table by case definition
df_symptoms_long |>
  group_by(case_def, symptom) |>
  summarize(reported = sum(symptom_experienced),
            percent_reported = scales::percent(reported/n())) |>
  pivot_wider(names_from = "case_def",
              values_from = c("reported", "percent_reported"))


# Table of cases
table_overview <- df_joined |> 
  group_by(region, )
  
  
  group_by(region) |> 
  summarize(n_total = n(),
            n_confirmed = sum(case_def=="Confirmed"),
            n_probable = sum(case_def == "Probable"),
            n_total_y
            earliest_report = min(date_report, na.rm=T),
            latest_report = max(date_report, na.rm=T))) |> 
  adorn_totals() ### CHANGE THIS




```
