---
title: "Introduction to R for<br>Applied Epidemiology"
subtitle: "COVID-19 case study"
author: "March 2022"
date: '[contact@appliedepi.org](mailto:contact@appliedepi.org)'
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: xaringan-themer.css
params:
  lang: "en"
---
class: center, middle, inverse, title-slide

```{r, eval=F, echo=F, include=F}
# Must do in order to render.

pacman::p_load(xaringan)
devtools::install_github("gadenbuie/xaringanExtra")
remotes::install_github("mitchelloharawild/icons")
icons::download_fontawesome()

# Render with xaringan::infinite_moon_reader()
# Slides will appear in viewer, and will update as you edit/save
```


```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3  ## retina more effective than w/h (according to twitter)
                      # fig.width = 16, 
                      # fig.height = 10
                      )
## I dont know why this is included in the example xaringan slide 
## but is something to do with background images
options(htmltools.dir.version = FALSE)

## install and load necessary packages 
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  sf,         # spatial
  ggspatial,  # spatial
  kableExtra, # for output tables
  xaringanthemer,  # for styling presentation
  tidyverse  # data management and visualization

)

## load packages from github
pacman::p_load_gh(
     "R4IDSR/epichecks"   # bivariate colour maps 
)


# data prep --------------------------------------------------------------------
surv <- rio::import(here::here("data", "surveillance_linelist_clean_20141201.rds"))

gen <- googlesheets4::read_sheet(
  "https://docs.google.com/spreadsheets/d/1-Xqv5xvakmhKSxipVP6a9GF3H54RN-rbilhdNhPHx6M/edit#gid=998182281",
  range = "generic") %>%
  clean_names()

mod10 <- googlesheets4::read_sheet(
  "https://docs.google.com/spreadsheets/d/1-Xqv5xvakmhKSxipVP6a9GF3H54RN-rbilhdNhPHx6M/edit#gid=998182281",
  range = "mod10", col_types = "c") %>%
  clean_names()
```

```{r  xaringan-themer, include = FALSE}

## define presentation colours (theme) using {xaringanthemer} package 
## https://pkg.garrickadenbuie.com/xaringanthemer/articles/xaringanthemer.html

## epirhandbook logo colours: 
  ## blue: "#00538c"
  ## green: "#007732"
  ## lighter green: "#48a878"

## see ?style_mono_accent for all the things can customise
style_mono_accent(
  base_color = "#00538c", 
  link_color = "#48a878", 
  ## add logo to the title page (bit bigger)
  title_slide_background_image = "https://github.com/appliedepi/intro_course/raw/main/images/logo.png", 
  title_slide_background_position = "95% 95%", 
  title_slide_background_size = "25%", 
  ## add logo to all following slides
  background_image = "https://github.com/appliedepi/intro_course/raw/main/images/logo.png", 
  background_size = "10%",
  background_position = "100% 0%"
)
```

```{css, echo=F}
    .remark-slide table{
      border: none
    }
    .remark-slide-table {
      
    }
    tr:first-child {
      border-top: none;
  }
    tr:last-child {
    border-bottom: none;
  }
```

# `r paste0(gen[gen$uid == "gen_h1_courseTitle", params$lang])`

### `r paste0(mod10[mod10$uid == "mod10_h1_subtitle", params$lang])`

contact@appliedepi.org
---

# `r paste0(mod10[mod10$uid == "mod10_h1_objectives", params$lang])`  

`r paste0(mod10[mod10$uid == "mod10_op_objectives", params$lang])`  

* `r paste0(mod10[mod10$uid == "mod10_bu_objectives_1", params$lang])`  

--

* `r paste0(mod10[mod10$uid == "mod10_bu_objectives_2", params$lang])`  

--

* `r paste0(mod10[mod10$uid == "mod10_bu_objectives_3", params$lang])`  

--

* `r paste0(mod10[mod10$uid == "mod10_bu_objectives_4", params$lang])`  

--

* `r paste0(mod10[mod10$uid == "mod10_bu_objectives_5", params$lang])`  

```{r, echo=FALSE, warning=F, message=F}
outline <- dplyr::tribble(
  ~Time, ~Location, ~Topic,
  str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_1_1", params$lang])}),     str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_2_1", params$lang])}),      str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_3_1", params$lang])}),
  str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_1_2", params$lang])}),        str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_2_2", params$lang])}),str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_3_2", params$lang])}),
  str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_1_3", params$lang])}),     str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_2_3", params$lang])}),      str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_3_3", params$lang])}),
  str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_1_4", params$lang])}),      str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_2_4", params$lang])}),      str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_3_4", params$lang])}),
  str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_1_5", params$lang])}),     str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_2_5", params$lang])}),      str_glue({paste0(mod10[mod10$uid == "mod10_tb_objectives_3_5", params$lang])})
)

outline %>% 
  flextable::qflextable() %>% 
  flextable::add_footer_lines(str_glue({paste0(mod10[mod10$uid == "mod10_fn_objectives", params$lang])}))
```


---

# `r paste0(mod10[mod10$uid == "mod10_h1_report", params$lang])`

`r paste0(mod10[mod10$uid == "mod10_tx_report_1", params$lang])`

`r paste0(mod10[mod10$uid == "mod10_tx_report_2", params$lang])`  


---

# `r paste0(mod10[mod10$uid == "mod10_h1_data", params$lang])`

`r paste0(mod10[mod10$uid == "mod10_tx_data_1", params$lang])`  

- `r paste0(mod10[mod10$uid == "mod10_bu_data_1", params$lang])`  
- `r paste0(mod10[mod10$uid == "mod10_bu_data_2", params$lang])`  


---

# `r paste0(mod10[mod10$uid == "mod10_h1_support", params$lang])`  

- `r paste0(mod10[mod10$uid == "mod10_bu_support_1", params$lang])`  
- `r paste0(mod10[mod10$uid == "mod10_bu_support_2", params$lang])`  


---
# `r paste0(mod10[mod10$uid == "mod10_h1_fun", params$lang])`  

`r paste0(mod10[mod10$uid == "mod10_tx_fun", params$lang])`  


---
# `r paste0(mod10[mod10$uid == "mod10_h1_review", params$lang])`  

`r paste0(mod10[mod10$uid == "mod10_tx_review_1", params$lang])`  

`r paste0(mod10[mod10$uid == "mod10_tx_review_2", params$lang])`  


---
# `r paste0(mod10[mod10$uid == "mod10_h1_feedback", params$lang])`  

`r paste0(mod10[mod10$uid == "mod10_tx_feedback", params$lang])`  

