---
title: "Module 5"
subtitle: "Instructor guide"
output: html_document
theme: sandstone
date: 'Date revised: 2022-03-04'
---


```{r, echo=FALSE}
# Put logo in upper-right of document
htmltools::img(src = knitr::image_uri(here::here("images", "logo.png")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width = "200px",
               heigth = "200px")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Notes  

The purpose of this demonstration is to show classic data transformations, which students will often encounter during their work: joining data and restructuring them. We will then have a short explanation on how to create and manipulate factors in R.  


# Outline  TODO

## Joins
### Slides + in-slide demo
- Introduce joining  
- Introduce and demonstrate each type of joins (left, right, inner, full)  
- Introduce and demonstrate filtering joins (semi and anti)  

### Demo
- Demonstrate with surveillance / medical linelsit  


### Break out
- Clarify any questions  
- Exercice: join together the surveillance linelist, medical linelist, spatial (district) locations, GIS coordinates, and source investigation dataset  


## BREAK

## Pivoting data

### Slides + in-slide demo
- Introduce long and large data format  
- Introduce and demonstrate pivoting  

### Break ouk
- Clarify question  
- Exercice  

## BREAK

## Factors
### Slides + in-slide demo
- Introduce concept of factors  



# Preparation  

1. Clear your desktop of clutter.  
2. Turn off pop-up notifications from desktop versions of WhatsApp, Signal, Teams, etc.  
3. Open RStudio and adjust the upper-right R project drop-down menu to close any open projects. This way, when you open RStudio it opens as "Project: None".  
4. Open a file explorer window, showing the location where you will save the R project, and have it ready to show.  
5. Open this instructor guide (or better: print it and have on your desk)

# Setup  

1. **In Google Meets**:  
     * If you have two displays: share your second display, which contains RStudio and the file explorer. The first display will be contain the Meet.  
     * If you only have one display, show your entire screen and ask an assistant instructor for help with any questions/hands that you cannot see. (TODO maybe there is a way to avoid mirroring in Meets?)  

</br>



# Joins

**A note on vocabulary** 
People coming from different background might have different words for this type of operation: joining, union, merging, combining, assembling...
In this module, we will use the word "join" to stick to the `join_` function familly from tidyverse and functions from SQL. The base R equivalent is `merge`.

## Objectives

**Learn how to join 2 or more tables together**  

More specifically:  
- be able to choose between the different sort of joins  
- be able to chose which to use ad keys/identifier  
- be able to troubleshoot the operation/perform it safely  





# Pivoting data

## Objectives

# Facors

## Objectives