---
title: "Applied Epi"
subtitle: "Taking control of your data"
format:
  revealjs: 
    code-line-numbers: false
    slide-number: true
    chalkboard: 
      buttons: true
    controls: false
    multiplex: false
    preview-links: auto
    logo: ../../images/Applied_Epi_logo.png
    css: ../../images/styles.css
    footer: <https://appliedepi.org>
resources:
  - demo.pdf
fig-align: center
---

```{r}

# Packages
pacman::p_load(rio, here, janitor, dplyr, gtExtras, gt, tidyr)

# Load data as needed 
df_raw <- import(here("intror/GIoutbreak_2021-06-05_linelist.csv"), na = "") 
lab_raw <- import(here("intror/GIoutbreak_2021-06-05_labs.xlsx"))

df_raw[df_raw == ""] <- NA

# Create the sub versions needed (df = as clean as appropriate for this module, _demo = filtered to fewer rows when need to show on slides)

df_raw_demo <- df_raw |> 
  filter(row_number()<6)

df <- df_raw %>% clean_names() |> rename(date_onset = symptoms_started)

df_demo <- df_raw_demo %>% clean_names() |> rename(date_onset = symptoms_started)

```

# Module 5: Part A

## Learning objectives

# Recap

# Concept

## Function

::: slide-header
{packagename}
:::

- *Purpose*: 
- *Why*: 
- *Example*: 
- *How*: 

::: {.code-box}
`code`
:::

## Example 1


::: slide-header
{dplyr}
:::

<br>

::: {.code-box-small style="width: 700px; font-size: 40px;"}
```{r echo=T, eval=F}
#Code goes here
```
:::

<br>

::: columns
::: {.column width="47%"}
::: input-box
Input
```{r}
#Code
```
:::

:::

::: {.column width="6%"}
::: arrow-box
`r r2symbols::symbol(3)`
:::
:::

::: {.column width="47%"}
::: output-box
Output
```{r}
#Code
```
:::
:::
:::


## Example 2

::: slide-header
{package}
:::

::: {.code-box}
`code`
:::

::: {.arrow-box}
&#8595;
:::


```{r}
glimpse(df_raw)
```

::: {.obs-box}
⚠️ **NOTE!** You could also write `data %>% glimpse()`. However as you will not continue the pipe chain, it is simpler to not use the pipe. 
:::
    
# Exercise

## Scenario {.exercise-slide}

::: incremental

- Point 1
- Point 2

:::

## Access the exercises {.exercise-slide}

Instructions here

# Recap

## Put code altogether

## Learning objectives
