---
title: "Functions and importing data"
subtitle: "Instructor guide"
output: html_document
date: 'Date revised: 2022-03-02'
---

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri(f = here::here("images", "Applied-Epi-logo.png")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Notes  

The purpose of this demonstration is to show how **base** R functions work, and then to introduce R packages, and finally to . We choose very simple functions that are commonly used in epidemiology.  

# Outline  

1. Introduce simple **base** R functions  
2. Introduce named arguments (TODO) 
3. Demonstrate installing a package  
4. Install **pacman* and use it to install other packages  
5. Import a dataset from the R project root folder and from a sub-folder  


# Preparation  

1. Clear your desktop of clutter.  
2. Turn off pop-up notifications from desktop versions of WhatsApp, Signal, Teams, etc.  
3. Open RStudio and adjust the upper-right R project drop-down menu to close any open projects. This way, when you open RStudio it opens as "Project: None".  
4. Open a file explorer window, showing the location where you will save the R project, and have it ready to show.  
5. Open this instructor guide (or better: print it and have on your desk)

# Setup  

1. **In Google Meets**:  
     * If you have two displays: share your second display, which contains RStudio and the file explorer. The first display will be contain the Meet.  
     * If you only have one display, show your entire screen and ask an assistant instructor for help with any questions/hands that you cannot see. (TODO maybe there is a way to avoid mirroring in Meets?)  

</br>

# Icons  

1. **Show desktop**  
     * Note the desktop RStudio icon, and that there is (or is not) a separate icon for R  

# Tour of RStudio  
1. **R projects**
     * **Adjust the zoom** so the text is readable by the audience. Confirm with one of the assistants.  
          * Note the **R project icon/drop-down in the upper-right**, which should say "Project: (None)".  
          * Note that one can begin working without a project, but we recommend using R projects  
     * **Open a new R project** (click through to standard and save in a **NEW directory**).  
          * **Save it** as "*intro_course*" in a pre-prepared location near where you have saved the course files.  
          * **Show the file explorer** and the new R project folder, and the .Rproj file.  
     * **Close RStudio, and re-open the project** by clicking the .Rproj file.  
          * Note the name of the project appears in the upper-right of RStudio. 
     * **Open a new R script**.  
          * Explain that this is where you will write code commands.  
          * Save it as "*demo.R*".  
          * Show the file appearing in the **Files pane**, and show it also in the File Explorer.  

</br>

2. **Tour the RStudio panes:**  
     * **Source Pane (R Script)**:  
          * R scripts ("instructions") - most commands are run/executed from here
          * Also used to view datasets  
          * For Stata-users, its like Do-file and Data Editor windows  
     * **R Console Pane**:  
          * The R "engine" - note the R version and the "common name" e.g. "Bird Hippie"  
          * You can run commands here too (but this is less common)  
          * Non-graphic outputs, errors, and warnings  
          * For Stata-users, its like the Command Window and also the Results window  
     * **Environment Pane**:  
          * Shows a list of the datasets you have imported or other objects you have created  
          * Briefly note History sub-pane  
          * For Stata-users, this is similar to the Variables Manager window  
     * **Files, Plots, Packages, Help, and Viewer**  
          * Open and rename files  
          * View graphic outputs  
          * Review your packages (*they won't know what packages are yet*)  
          * See help documentation  
          * For Stata-users this contains Plots and Project Manger windows  

</br>

# Basic R syntax  

1. **R commands**  
     * Begin typing into the R script  
     * Type `1 + 2`, note that your cursor is in this command line, and click "Run" with your mouse.  
          * Show the result appear in the R Console  
          * Note the `>` that shows the original command  
          * Note the `[1]` that precedes the answer. This can confuse some people.
     * Cut & Paste the same command to a different line in the script. Run it again.  
     * Add/change the command to make `(1 + 2) * 4 / 6`.  
          * Ask the audience what the answer will be (put in the chat)  
          * Run the entire command and see the answer `2`  
          * Highlight just the segment `1 + 2` and click Run.  
          * Note that only the highlighted portion was executed. 
     * Explain that you can run the command by placing your cursor anywhere in the line and typing *Ctrl and Enter* (Cmd and Enter on macs)  

</br>

2. **Define numeric object**  
     * Explain that you are going to define an object, and introduce R's "assignment operator" `<-`.  
     * Explain that it takes the **value** of the right side, and assigns it to the **name** on the left.  
     * The operator can be used to **create** an object, or to re-assign its value:  
          * In the script, type and run `epiweek <- 34`. Explain that you are creating an object named "epiweek" that has the numeric value 34.  
          * Explain that for anyone unfamiliar, an "epiweek" is shorthand for an "epidemiological week" and helps standardise reporting.  
          * Show how this new object appears in the **Environment Pane**.
          * Type the command `epiweek` in the R console and see the value `34` appear below  
          * Highlight the word "epiweek" *in the R script* and run it. See the value `34` appear in the Console.  

</br>

3. **Use a defined numeric value**  
     * Write the command `epiweek - 1` in the script.  
          * Ask the audience what the value will be (enter in the chat).  
          * Run the command and note the output `33`.  
          * Write the command `previous_week <- epiweek - 1`.  
          * Briefly note that objects cannot have spaces, and cannot start with a number  
          * Ask the audience where changes will appear in RStudio when the command is run.  
          * Run it, and note Console only shows record of the command - it does not print a value in the Console (note lack of `[1]`). Note new object name and value appears in Environment Pane.  
          * Note the difference between commands that *print* a object or value, and commands that *define* an object (**"What are you asking R to do?"**)  
          * Run the command `previous_week` from the script. Note the printed value `33`.  

</br>

3. **Re-define the value**  
     * *Write a new command* below the old command `epiweek <- 35` in the script, and run it.  
          * Note the new value in the **Environment Pane**  
          * Note that `previous_week` is still `33`. Re-run `previous_week <- epiweek - 1` and note the change.  
     * **Edit** the previous command to read `epiweek <- 36` and run it. Note the new value in the Environment Pane.  
          * Run just the word `epiweek` from the script. Note the value 36.  

# Commenting  

1. **Introduce script header**  
     * * Explain that now that we have several commands, we should organize it. For ourselves and any future readers.  
     * Begin by writing a few lines at the top:  
     
```{r, eval=F, echo=T}
# Demo script for intro course
##############################
# Last revised: DATE
# Written by: YOUR NAME
# Contact: contact@appliedepi.org
```
   
2. **Introduce line-by-line comments**  
     * Explain that everyone has their own particular style, but this is one option.  
     * Explain that readability is important - give lots of *vertical* space and write clearly.  
     
```{r, eval=F, echo=T}
# Define epiweek  
epiweek <- 36

# Define previous epiweek  
previous_week <- epiweek - 1
```

3. **Right-side comments**  
     * Explain that nothing to the right of the # will be run.  
     * Re-write your comments to appear *after* and *to the right* of each command:  

```{r, eval=F, echo=T}
epiweek <- 36                  # define epiweek  
previous_week <- epiweek - 1   # define previous epiweek  
```

# Another example  

1. Give another example of defining objects
     * Write the below, step-by step. Run each command sequentially.  

```{r, eval=F, echo=T}
# Suspect cases
suspect_cases <- 2    # define
suspect_cases         # print

# Confirmed cases  
confirmed_cases <- 10 # define
confirmed_cases       # print

# Total cases
total_cases <- suspect_cases + confirmed_cases   # calculate
total_cases # print
```

# Other class of objects    
1. Define a character object  
     * Introduce the idea of *classes* - but do not introduce `class()` yet.
     * Note that until now we have dealt with numbers, but you can also define "character values"
     * Write the command `district <- "Bolo"` in the script and run it.  
     * Print by running the command `district`.
     * Note the difference between `"34"` and `34` - one you can use as a number, the other you should not.    
     
# Course files  
1. **Copy course files into the R project**  
     * Toggle out of RStudio and show the File Explorer window.
          * **Copy the course files** and paste them into the R project  
          * Show how they are accessible from the RStudio **Files pane**  
          * Discuss the structure (a "data" folder, a "scripts" folder, and a "outputs" folder)          
     * Save and close "*demo.R*" in RStudio. In the File Explorer, move it into the "scripts" folder manually.  

# Finish
Close RStudio.    



     
     
     


</br>

</br>

</br>

</br>
